### 2014-08-29 paper/paper.Rmd
updated with luke's code and comments

student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
  scale_fill_discrete("Median Start Age") +
  new_theme_empty + theme(legend.position="bottom")
#ggsave("agestartmap.pdf", width=12, height=7)
```
### Time spent out of school studying is important, but only up to a point

***Need to get this code working

```{r, eval=FALSE}
# Out of School Study time
b1 <- summaryBy(data=student2012, pvM + pvR + pvS+ ST57Q01+ ST57Q02 +ST57Q03+ ST57Q04+ ST57Q05+ ST57Q06 ~CNT,FUN=mean, na.rm=T)
names(b1) <- c("CNT","pvM","pvR","pvS","ST57Q01", "ST57Q02","ST57Q03", "ST57Q04", "ST57Q05", "ST57Q06" )
# Out of school study time (total hours HW)
b1$CNT <- reorder(factor(b1$CNT),b1$pvS, mean)
b2 <- melt(b1[,c("CNT","pvM","pvR","pvS")],id="CNT")
b2 <- cbind(b2, b1[match(b2$CNT,b1$CNT), 5:10])
qplot(CNT, value,color=variable,size=ST57Q01, data = b2)+ coord_flip()
head(b2)
qplot(ST57Q01, value,color=variable,size=ST57Q02, data = b2)+ coord_flip()+facet_wrap(~CNT)

student.sub <- studschool[,c("name","pvM","pvR","pvS","ST57Q01","OECD","SC03Q01")]
study <- melt(student.sub,id=c("name","ST57Q01","OECD","SC03Q01"))
names(study) <- c("Country","HomeworkHours","OECD","AREA","Subject","Score")
levels(study$Subject) <- c("Math","Reading","Science")
p <- ggplot(data = study, aes(HomeworkHours,Score,colour=Subject) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)+ xlab("Out-of-School Study Hours")
#ggsave(p,file="studytime.pdf",height=13,width=13)
p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") +xlim(0,30)
ggsave(p,file="HistHWschool.pdf",width=8,height=6)

student.sub <- studschool[,c("name","pvM","ST57Q01","SC03Q01")]
study <- student.sub
names(study) <- c("Country","Math","HomeworkHours","Schoolplace")

p <- ggplot(data = study, aes(HomeworkHours,Math,colour=Schoolplace) )+xlim(0,20)+facet_wrap(~Country)+stat_smooth(se=F)
#ggsave(p,file="studytime.pdf",height=13,width=13)
p <- ggplot(data=student.sub, aes(ST57Q01, ..density..)) + geom_histogram(binwidth=5) + xlab("Out-of-School Study Hours") + facet_grid(SC03Q01~.) +xlim(0,30)
```

### Parents matter

The role of parents in a child's education and success later in life is of great interest to both educators and policy makers. The relationship between eduational achievement and both household structure and parent's occupational status were evaluated. To look at the relationship between educational success and family structure, boxplots were made of the student's scores conditioned whether the student had a two-parent household, only a mother at home, only a father at home, or no parents at home. The relationship was evaluated for every participating country. The majority of the students did live in a two-parent household, but there were still sizable numbers of students who did not. The information regarding a parent's occupational was provided in a way that can be broadly catergorized with five catergories: "Full-employment", "Part-time", "Unemployed-looking for work", "Other: Home duties, retired, etc.", and "Missing". The means achievement in mathematics for each of the categories was calculated and plotted for each country.

```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
athome = student2012[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = filter(athome, !is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
#athome$name <- reorder(athome$name, athome$PV1MATH, mean, decreasing=F)
#qplot(x = parents, data = athome) 
```

```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math Scores") + xlab("Family Structure") 
```

One of the variables the survey addressed was the occupation of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
#student2012 <- student2012
student2012$ST15Q01 <- addNA(student2012$ST15Q01)
student2012$ST19Q01 <- addNA(student2012$ST19Q01)

#student2012$name <- as.character(student2012$CNT)
# unique(anti_join(student2012.sub, world.polys)[1])
#student2012$name[student2012$name=="United Arab Emirates"] <- "UAE"
#student2012$name[student2012$name=="United Kingdom"] <- "UK"
#student2012$name[student2012$name=="Serbia"] <- "Serbia"
#student2012$name[student2012$name=="Korea"] <- "South Korea"
#student2012$name[student2012$name=="Chinese Taipei"] <- "Taiwan"
#student2012$name[student2012$name=="Slovak Republic"] <- "Slovakia"
#student2012$name[student2012$name=="Russian Federation"] <- "Russia"
#student2012$name[student2012$name=="Perm(Russian Federation)"] <- "Russia"
#student2012$name[student2012$name=="Hong Kong-China"] <- "Hong Kong"
#student2012$name[student2012$name=="China-Shanghai"] <- "China"
#student2012$name[student2012$name=="Macau"] <- "China"
#student2012$name[student2012$name=="Connecticut (USA)"] <- "USA"
#student2012$name[student2012$name=="Florida (USA)"] <- "USA"
#student2012$name[student2012$name=="Massachusetts (USA)"] <- "USA"
#student2012$name[student2012$name=="United States of America"] <- "USA"
#unique(student2012$name)
#student2012$name <- factor(student2012$name)

a6 <- summaryBy(data=student2012, PV1MATH + ESCS ~ ST15Q01+name, FUN=mean, rm.na=T)
a7 <- summaryBy(data=student2012, PV1MATH + ESCS ~ ST19Q01+name, FUN=mean, rm.na=T)
colnames(a6) <- colnames(a7) <- c("Job.Status","name","Math","ESCS")
levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")

a8 <- rbind(cbind(a6, Parent="Mother"), cbind(a7, Parent="Father"))
#a2 <- melt(a1)
ind <- which(a8$Job.Status=="Other" & a8$Parent=="Mother")
a8$name <- factor(a8$name, a8$name[order(a8$Math[ind])])
qplot(name, Math, col=Job.Status, data = a8) + coord_flip() + facet_wrap(~Parent)
```

>>>>

student2012.sub.agemedian$medianstart <- factor(student2012.sub.agemedian$medianstart)
student2012.sub.map <- left_join(student2012.sub.agemedian, world.polys)
ggplot(data=world.polys) + geom_path(aes(x=X1, y=X2, order=order, group=group), colour=I("grey90")) + 
  geom_polygon(data=student2012.sub.map, aes(x=X1, y=X2, order=order, group=group, fill=medianstart)) +
  scale_fill_discrete("Median Start Age") +
  new_theme_empty + theme(legend.position="bottom")
#ggsave("agestartmap.pdf", width=12, height=7)
```
### Time spent out of school studying is important, but only up to a point
### Time spent out of school studying is important, but only up to a point
For each country we have provided loess smoothed curves of the relationship between mathematics and the time spent studying outside of school. The most prominent feature that we observe is the decreasing benefit of studying. For every country there is evidence of a substantial positive relationship for the students who spend less time studying. However, those students who continue to study for long hours see diminishing returns for their efforts. It is possible that there may be a selection issue occuring where more talented students spend less time studying as they complete their homework faster. However, the relationship shows those spending

```{r studying, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
student.luke <- student2012[,c("CNT","PV1MATH","PV1READ","PV1SCIE","ST57Q01")]
study.luke <- melt(student.luke,id=c("CNT","ST57Q01"))
studyplot <- ggplot(data = study.luke, aes(ST57Q01,value,colour=variable) )+facet_wrap(~CNT)+stat_smooth(se=F)
studyplot
```

```{r studyingbar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
studybar <- ggplot(data = study.luke, aes(ST57Q01) ) + geom_histogram(binwidth=4) + xlab("Hours spent studying")
studybar
```

### Parents matter
### subsection: Number of parents in the home
Looking at the number of parents in a household shows that student in two-parent household are on average performing better than student's not in a two-parent household. Students with just a mother at home do tend to perform better than students with only a father at home. Students with no parents at home are performing the worst, on average.

```{r par_athomebar, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
athome = studentLuke[,c("name", "SCHOOLID", "STIDSTD", "OECD", "PV1MATH", "PV1SCIE", "PV1READ","ST11Q01", "ST11Q02", "ST11Q03","ST11Q04", "ST11Q05", "ST11Q06")]
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "Both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "Mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "Father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "Neither"
athome$name <- reorder(athome$name,athome$PV1MATH,mean)
pbarhome <- qplot(x = parents, data = athome) 
pbarhome

```{r par_athome, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
phome <- qplot(x = parents, y = PV1MATH, geom = "boxplot", outlier.size = 1, facets = ~name, data = athome) + coord_flip() + ylab("Math") + xlab("Family Structure") 
phome
```

### subsection: Parent's Occupation Status
One of the variables the survey addressed was the occupation status of each of the parents. There were five possible categories: “Full-Time”, “Part-Time”, “Not working, but looking for a job”, “Other (e.g. home duties, retired)”, and no response. The distribution of the number of parents who selected each of the categories can be seen in the bar chart. The chart shows that father’s primarily have full-time jobs. However, there are large numbers of mothers who work full-time, part-time, or are stay-at-home moms. With women making up a larger portion of the work-force, it is interesting to look at how the parent’s occupation relates to the student’s performance in mathematics, reading, and science.

```{r parentsjob, fig.width=10, fig.height=12, echo = FALSE, message = FALSE, warning = FALSE}
studentLuke <- student2012
studentLuke$ST15Q01 <- addNA(studentLuke$ST15Q01)
studentLuke$ST19Q01 <- addNA(studentLuke$ST19Q01)

studentLuke$name <- as.character(studentLuke$CNT)
# unique(anti_join(studentLuke.sub, world.polys)[1])
studentLuke$name[studentLuke$name=="United Arab Emirates"] <- "UAE"
studentLuke$name[studentLuke$name=="United Kingdom"] <- "UK"
studentLuke$name[studentLuke$name=="Serbia"] <- "Serbia"
studentLuke$name[studentLuke$name=="Korea"] <- "South Korea"
studentLuke$name[studentLuke$name=="Chinese Taipei"] <- "Taiwan"
studentLuke$name[studentLuke$name=="Slovak Republic"] <- "Slovakia"
studentLuke$name[studentLuke$name=="Russian Federation"] <- "Russia"
studentLuke$name[studentLuke$name=="Perm(Russian Federation)"] <- "Russia"
studentLuke$name[studentLuke$name=="Hong Kong-China"] <- "Hong Kong"
studentLuke$name[studentLuke$name=="China-Shanghai"] <- "China"
studentLuke$name[studentLuke$name=="Macau"] <- "China"
studentLuke$name[studentLuke$name=="Connecticut (USA)"] <- "USA"
studentLuke$name[studentLuke$name=="Florida (USA)"] <- "USA"
studentLuke$name[studentLuke$name=="Massachusetts (USA)"] <- "USA"
studentLuke$name[studentLuke$name=="United States of America"] <- "USA"
unique(studentLuke$name)
studentLuke$name <- factor(studentLuke$name)

a6 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST15Q01+ name,FUN=mean,rm.na=T)
a7 <- summaryBy(data=studentLuke, PV1MATH + ESCS ~ ST19Q01+ name,FUN=mean,rm.na=T)
colnames(a6) <- colnames(a7) <- c("Job.Status","Country","Math","ESCS")
levels(a6$Job.Status) <- levels(a7$Job.Status)  <- c("Full-Time","Part-Time","Unemployed","Other","NA")

a8 <- rbind(cbind(a6,Parent="Mother"),cbind(a7,Parent="Father"))
#a2 <- melt(a1)
ind <- which(a8$Job.Status=="Other"& a8$Parent=="Mother")
a8$Country <- factor(a8$Country,levels(a8$Country)[order(a8$Math[ind])])
p <- qplot(Country, Math, col=Job.Status, data = a8) +coord_flip() + facet_wrap(~Parent)
p
```

The Figure ?*? shows the average performance of student’s performance based on the parent’s occupational status. The countries are sorted according to the average score for students with a "stay-at-home" mother. The averages for each parent in each of the work status categories are distinguished by color. As you can see in each of the plots below, students’ with a mother who works part-time or full-time tend outperform those students with a stay-at-home mother. This pattern holds with only a few exceptions. Meanwhile, when the father is employed part-time students tend to have much lower achievement that if the father stayed at home. When the father has a full-time jobs, the student’s have the highest average achievement among the possible work status’ for the father. The patterns described hold across mathematics, reading, and science.
